---
title: "p8105_hw2_ajt2206"
author: "Andy Turner"
date: "2023-09-29"
output: github_document
---
**Libraries**
```{r}
library(tidyverse)
```



## Problem 1

Importing data into R
```{r}
pols_df=
  read_csv("data/pols-month.csv") |> 
  janitor::clean_names() |> 
  separate(mon, into=c("year", "month", "day"), sep= "-") |> 
  mutate(
    month= case_match(
      month,
      "01" ~ "January",
      "02" ~ "February",
      "03" ~ "March",
      "04" ~ "April",
      "05" ~ "May",
      "06" ~ "June",
      "07" ~ "July", 
      "08" ~ "August",
      "09" ~ "September",
      "10"~ "October",
      "11"~ "November",
      "12"~ "December"
    )
  ) |> 
  mutate(
    prez_dem= case_match(
      prez_dem,
      0 ~ "GOP",
      1 ~ "DEM"
    )
  ) |> 
  mutate(president = prez_dem) |> 
  select(-prez_gop, -prez_dem, -day)
 
```

## Problem 2

**Reading and Cleaning the Mr. Trash Wheel data**

Prior to importing the Mr. Trash Wheel data into R, I checked the excel file. The "Year" column had an error where the numeric year was stored as text. I converted the column data to numeric format within Excel (Professor and Gwynnda already had these values stored as numeric). 

```{r}
mr_trash_df=
  readxl::read_excel("data/202207 Trash Wheel Collection Data.xlsx", sheet= "Mr. Trash Wheel", range="A2:N549") |> 
  janitor::clean_names() |> 
  mutate(homes_powered= weight_tons*500/30) |>
  mutate(homes_powered= round(homes_powered)) |> 
  mutate(wheel= "Mr. Trash Wheel") |> 
  select(wheel, everything())
```
To read-in and clean up the Mr. Trash Wheel data set, I performed the following steps:

1. Used the `readxl` to read in the data, restricting the sheet and rows to only Mr. Trash Wheel with data. 

1. `clean_names()` was used to make sure all the imported variables were consistently named

1. I overwrote existing values in homes_powered using the equation found in Homes Powered Note in the original excel file. 

1. Because we are looking for an approximation of homes_powered, I rounded the values in homes_powered to the nearest whole number make the data more easily readable. 

1. I created a new variable "wheel" to help identify each trash wheel's data once combined. 

1. Last, I re-ordered the data frame slightly to put the wheel variable first to simply make the rows more identifiable once combined as well. 

**Reading and Cleaning the Professor and Gwynnda Trash Wheel Data**
```{r}
prof_trash_df=
  readxl::read_excel("data/202207 Trash Wheel Collection Data.xlsx", sheet= "Professor Trash Wheel", range="A2:M96") |> 
  janitor::clean_names() |> 
  mutate(homes_powered= weight_tons*500/30) |> 
  mutate(homes_powered= round(homes_powered)) |>
  mutate(wheel= "Professor Trash Wheel") |> 
 select(wheel, everything())

gwynnda_trash_df=
  readxl::read_excel("data/202207 Trash Wheel Collection Data.xlsx", sheet= "Gwynnda Trash Wheel", range="A2:K108") |> 
  janitor::clean_names() |> 
  mutate(homes_powered= weight_tons*500/30) |> 
  mutate(homes_powered= round(homes_powered)) |>
  mutate(wheel= "Gwynnda Trash Wheel") |> 
 select(wheel, everything())
```
Same steps as described above for Mr. Trash Wheel data were repeated for the Professor and Gwynnda Trash Wheels. 

**Combining the three datasets**
```{r}
triple_trash_df=
  bind_rows(mr_trash_df, prof_trash_df, gwynnda_trash_df)
```

**Description of the Data**

* This dataset provides data from May 2014 to August 2018 on the trash collected by three Trash Wheels situated in the Inner Harbor in Baltimore, Maryland. 

* There are `r nrow(triple_trash_df)` observations and `r ncol(triple_trash_df)` variables within this dataset. 

* The variables are as follows `r names(triple_trash_df)`. 

* Key variables include the following. `wheel`: identifies which of the three trash wheels the row corresponds to. `weight_tons`: how much trash in tons in each dumpster. `homes_powered`: an estimation of how many Maryland homes are powered by the trash collected per dumpster. 

* Total weight of trash collected by Professor Trash Wheel is tons. `sum(prof_trash_df, weight_tons)`

* Total number of cigarette butts collected by Gwynnda is July of 2021: 

## Problem 3

```{r}
MCI_baseline_df=
  readxl::read_excel("data/MCI_baseline.xlsx", range="A2:F485") |> 
  janitor::clean_names() |> 
  mutate(
    sex= case_match(
      sex,
      0 ~ "Female",
      1 ~ "Male")
  ) |> 
  mutate(
   apoe4= case_match(
      apoe4,
      0 ~ "Non-carrier",
      1 ~ "Carrier")
  ) |> 
  filter(!(age_at_onset == "."))
```
**Description of the import process and dataset**

`r nrow(MCI_baseline_df)` participants met the inclusion criteria of the study and were included in the data.  

Discuss important steps in the import process and relevant features of the dataset. How many participants were recruited, and of these how many develop MCI? What is the average baseline age? What proportion of women in the study are APOE4 carriers?