---
title: "p8105_hw2_ajt2206"
author: "Andy Turner"
date: "2023-09-29"
output: github_document
---
**Libraries**
```{r}
library(tidyverse)
```



## Problem 1

Importing data into R
```{r}
pols_df=
  read_csv("data/pols-month.csv") |> 
  janitor::clean_names() |> 
  separate(mon, into=c("year", "month", "day"), sep= "-") |> 
  mutate(
    month= case_match(
      month,
      "01" ~ "January",
      "02" ~ "February",
      "03" ~ "March",
      "04" ~ "April",
      "05" ~ "May",
      "06" ~ "June",
      "07" ~ "July", 
      "08" ~ "August",
      "09" ~ "September",
      "10"~ "October",
      "11"~ "November",
      "12"~ "December"
    )
  ) |> 
  mutate(
    prez_dem= case_match(
      prez_dem,
      0 ~ "GOP",
      1 ~ "DEM"
    )
  ) |> 
  mutate(president = prez_dem) |> 
  select(-prez_gop, -prez_dem, -day)
 
```

## Problem 2

**Reading and Cleaning the Mr. Trash Wheel data**

Prior to importing the Mr. Trash Wheel data into R, I checked the excel file. The "Year" column had an error where the numeric year was stored as text. I converted the column data to numeric format within Excel (Professor and Gwynnda already had these values stored as numeric). 

```{r}
mr_trash_df=
  readxl::read_excel("data/202309 Trash Wheel Collection Data.xlsx", sheet= "Mr. Trash Wheel", range="A2:N586") |> 
  janitor::clean_names() |> 
  mutate(homes_powered= weight_tons*500/30) |>
  mutate(homes_powered= round(homes_powered)) |> 
  mutate(wheel= "Mr. Trash Wheel") |> 
  select(wheel, everything())
```
To read-in and clean up the Mr. Trash Wheel data set, I performed the following steps:

1. Used the `readxl` to read in the data, restricting the import to only include the Mr. Trash Wheel sheet and to columns/rows with data. 

1. `clean_names()` was used to make sure all the imported variables were consistently named. 

1. I overwrote existing values in homes_powered using the equation found in Homes Powered Note in the original excel file. 

1. Because we are looking for an approximation of homes_powered, I rounded the values in homes_powered to the nearest whole number make the data more easily readable using `round(0)`. 

1. I created a new variable "wheel" to help identify each trash wheel's data once combined. 

1. Last, I re-ordered the data frame slightly to put the wheel variable first to simply make the rows more identifiable once combined as well. 

**Reading and Cleaning the Professor and Gwynnda Trash Wheel Data**
```{r}
prof_trash_df=
  readxl::read_excel("data/202309 Trash Wheel Collection Data.xlsx", sheet= "Professor Trash Wheel", range="A2:M108") |> 
  janitor::clean_names() |> 
  mutate(homes_powered= weight_tons*500/30) |> 
  mutate(homes_powered= round(homes_powered)) |>
  mutate(wheel= "Professor Trash Wheel") |> 
 select(wheel, everything())

gwynnda_trash_df=
  readxl::read_excel("data/202309 Trash Wheel Collection Data.xlsx", sheet= "Gwynnda Trash Wheel", range="A2:L157") |> 
  janitor::clean_names() |> 
  mutate(homes_powered= weight_tons*500/30) |> 
  mutate(homes_powered= round(homes_powered)) |>
  mutate(wheel= "Gwynnda Trash Wheel") |> 
 select(wheel, everything())
```
Same steps as described above for Mr. Trash Wheel data were repeated for the Professor and Gwynnda Trash Wheels. 

**Combining the three datasets**
```{r}
triple_trash_df=
  bind_rows(mr_trash_df, prof_trash_df, gwynnda_trash_df)
```
To combine the datasets, I used the `bind_rows()` function. The datasets were essentially set on top on one another, but since we have the "wheel" variable, it is easy to tell the origin of each observation. 


**Description of the Data**

* This dataset provides data from May 2014 to August 2018 on the trash collected by three Trash Wheels situated in the Inner Harbor in Baltimore, Maryland. There are `r nrow(triple_trash_df)` observations and `r ncol(triple_trash_df)` variables within this dataset. The variables are as follows `r names(triple_trash_df)`. Key variables include the following. `wheel`: identifies which of the three trash wheels the row corresponds to. `weight_tons`: how much trash in tons in each dumpster. `homes_powered`: an estimation of how many Maryland homes are powered by the trash collected per dumpster. Professor Trash Wheel collected `r filter(triple_trash_df, wheel== "Professor Trash Wheel") |> pull(weight_tons) |> sum()`tons of trash as of September 2023. Moreover, Gwynnda Trash Wheel collected a total of `r filter(triple_trash_df, wheel== "Gwynnda Trash Wheel", month == "July", year == "2021") |> pull(cigarette_butts) |> sum()` cigarette butts during the month of July 2021. 

## Problem 3
```{r}
mci_baseline_df=
  readxl::read_excel("data/mci_baseline.xlsx", range="A2:F485") |> 
  janitor::clean_names() |> 
  mutate(
    sex= case_match(
      sex,
      0 ~ "Female",
      1 ~ "Male")
  ) |> 
  mutate(
   apoe4= case_match(
      apoe4,
      0 ~ "Non-carrier",
      1 ~ "Carrier")
  ) |> 
  filter(current_age != age_at_onset)
```
To read-in and clean up the MCI Baseline data set, I performed the following steps:

1. Used the `readxl` to read in the data, restricting the import to only include the columns and rows with data along with the variable names. 

1. `clean_names()` was used to make sure all the imported variables were consistently named. 

1. I changed "sex" from being stored as a binary numeric value to a character vector better identifying which participants are Male vs. Female. 

1. I changed "apoe4" from being stored as a binary numeric value to a character vector better identifying which participants are Carriers vs. Non-Carriers. 

1. As part of the study inclusion criteria, participants had to be free of MCI at study baseline. To make sure participants were free of MCI, I filtered the data to only include participants whose age at study baseline was different that their age of MCI onset. 


**Description of the import process and dataset**

`r nrow(mci_baseline_df)` participants met the inclusion criteria of the study and were included in the data. One participant was excluded as they had mci at Baseline. Of the `r nrow(mci_baseline_df)` participants, `r filter(mci_baseline_df, age_at_onset != ".") |> nrow()` went on to develop mci. The average baseline age is `r mci_baseline_df |> pull(current_age) |> mean()` years old. `r filter(mci_baseline_df, sex == "Female", apoe4 == "Carrier") |> nrow() / filter(mci_baseline_df, sex == "Female") |> nrow() * 100` % of the women in the study are APOE4 carriers. 

```{r}
mci_amyloid_df=
  readxl::read_excel("data/mci_amyloid.xlsx", range="A2:F489") |> 
  janitor::clean_names() |> 
  rename(id=study_id) |> 
  pivot_longer(
    baseline:time_8,
    names_to = "yrs_since_baseline",
    values_to = "amyloid_ratio",
    names_prefix= "time_"
  ) |> 
  mutate(
   yrs_since_baseline= replace(yrs_since_baseline, yrs_since_baseline== "baseline", "0")
  )

```
To read-in and clean up the MCI Amyloid data set, I performed the following steps:

1. Used the `readxl` to read in the data, restricting the import to only include the columns and rows with data along with the variable names. 

1. `clean_names()` was used to make sure all the imported variables were consistently named. 

1. I renamed the "study_id" variable to "id" to reflect the naming convention in the `mci_baseline_df` dataset.  

1. I changed "apoe4" from being stored as a binary numeric value to a character vector better identifying which participants are Carriers vs. Non-Carriers. 

1. Originally the data was in horizontal format, I decided to use `pivot_longer()` to switch it to vertical. The variable names (baseline: time_8) became values of the "yrs_since_baseline" variable, and the values were filled into the new "amyloid_ratio" variable. Additionally while pivoting, I removed the "time_" prefix.

1. I renamed "baseline" to "0" to ensure that the "yrs_since_baseline" variable had consistent value formatting throughout.

*Features of the dataset**

There are `r nrow(mci_amyloid_df)/5` participants in the dataset corresponding to `r nrow(mci_amyloid_df)` observations overall as each participant has 5 observations of data. There are `r ncol(mci_amyloid_df)` variables in the dataset. The variables are as follows: `r names(mci_amyloid_df)`. id: corresponds to participant id in the study; yrs_since_baseline: Time (in years) elapsed since the study baseline to the visit where biomarker Amyloid _ 42/40 ratio was measured; amyloid_ratio: the amyloid ratio measured

```{r}
base_unique = anti_join(mci_baseline_df, mci_amyloid_df) 

amyl_unique= anti_join(mci_amyloid_df,mci_baseline_df)|> 
  distinct(id, .keep_all= T)
```
**Unique Participants**

`r nrow(base_unique)` participants appear only only in the baseline dataset and `r nrow(amyl_unique)` participants appear only in the amyloid dataset.

Combine the demographic and biomarker datasets so that only participants who appear in both datasets are retained, and briefly describe the resulting dataset; export the result as a CSV to your data directory

```{r}
mci_combined_df=
  inner_join(mci_baseline_df, mci_amyloid_df, by= "id")
```
To combine datasets in a way in which only participants who appear in both datasets were retained, I used the `inner_join` feature and joined them by "id".

**Description of the combined dataset**
