---
title: "p8105_hw2_ajt2206"
author: "Andy Turner"
date: "2023-09-29"
output: github_document
---
**Libraries**
```{r}
library(tidyverse)
```



## Problem 1

Importing data into R
```{r}
pols_df=
  read_csv("data/pols-month.csv") |> 
  janitor::clean_names() |> 
  separate(mon, into=c("year", "month", "day"), sep= "-") |> 
  mutate(
    month= case_match(
      month,
      "01" ~ "January",
      "02" ~ "February",
      "03" ~ "March",
      "04" ~ "April",
      "05" ~ "May",
      "06" ~ "June",
      "07" ~ "July", 
      "08" ~ "August",
      "09" ~ "September",
      "10"~ "October",
      "11"~ "November",
      "12"~ "December"
    )
  ) |> 
  mutate(
    prez_dem= case_match(
      prez_dem,
      0 ~ "GOP",
      1 ~ "DEM"
    )
  ) |> 
  mutate(president = prez_dem) |> 
  select(-prez_gop, -prez_dem, -day)
 
```

## Problem 2

**Reading and Cleaning the Mr. Trash Wheel data**

Prior to importing the Mr. Trash Wheel data into R, I checked the excel file. The "Year" column had an error where the numeric year was stored as text. I converted the column data to numeric format within Excel (Professor and Gwynnda already had these values stored as numeric). 

```{r}
mr_trash_df=
  readxl::read_excel("data/202207 Trash Wheel Collection Data.xlsx", sheet= "Mr. Trash Wheel", range="A2:N549") |> 
  janitor::clean_names() |> 
  mutate(homes_powered= weight_tons*500/30) |>
  mutate(homes_powered= round(homes_powered)) |> 
  mutate(wheel= "Mr. Trash Wheel") |> 
  select(wheel, everything())
```
To read-in and clean up the Mr. Trash Wheel data set, I performed the following steps:

1. Used the `readxl` to read in the data, restricting the sheet and rows to only Mr. Trash Wheel with data. 

1. `clean_names()` was used to make sure all the imported variables were consistently named

1. I overwrote existing values in homes_powered using the equation found in Homes Powered Note in the original excel file. 

1. Because we are looking for an approximation of homes_powered, I rounded the values in homes_powered to the nearest whole number make the data more easily readable. 

1. I created a new variable "wheel" to help identify each trash wheel's data once combined. 

1. Last, I re-ordered the data frame slightly to put the wheel variable first to simply make the rows more identifiable once combined as well. 

**Reading and Cleaning the Professor and Gwynnda Trash Wheel Data**
```{r}
prof_trash_df=
  readxl::read_excel("data/202207 Trash Wheel Collection Data.xlsx", sheet= "Professor Trash Wheel", range="A2:M96") |> 
  janitor::clean_names() |> 
  mutate(homes_powered= weight_tons*500/30) |> 
  mutate(homes_powered= round(homes_powered)) |>
  mutate(wheel= "Professor Trash Wheel") |> 
 select(wheel, everything())

gwynnda_trash_df=
  readxl::read_excel("data/202207 Trash Wheel Collection Data.xlsx", sheet= "Gwynnda Trash Wheel", range="A2:K108") |> 
  janitor::clean_names() |> 
  mutate(homes_powered= weight_tons*500/30) |> 
  mutate(homes_powered= round(homes_powered)) |>
  mutate(wheel= "Gwynnda Trash Wheel") |> 
 select(wheel, everything())
```
Same steps as described above for Mr. Trash Wheel data were repeated for the Professor and Gwynnda Trash Wheels. 

**Combining the three datasets**
```{r}
triple_trash_df=
  bind_rows(mr_trash_df, prof_trash_df, gwynnda_trash_df)
```

**Description of the Data**

* This dataset provides data from May 2014 to August 2018 on the trash collected by three Trash Wheels situated in the Inner Harbor in Baltimore, Maryland. 

* There are `r nrow(triple_trash_df)` observations and `r ncol(triple_trash_df)` variables within this dataset. 

* The variables are as follows `r names(triple_trash_df)`. 

* Key variables include the following. `wheel`: identifies which of the three trash wheels the row corresponds to. `weight_tons`: how much trash in tons in each dumpster. `homes_powered`: an estimation of how many Maryland homes are powered by the trash collected per dumpster. 

* Total weight of trash collected by Professor Trash Wheel is tons. `sum(prof_trash_df, weight_tons)`

* Total number of cigarette butts collected by Gwynnda is July of 2021: 

## Problem 3


